---
title: "Response To Reviewer Mine "
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> This paper does a great job outlining some of the difficulties with working with categorical data, specifically factors, in R. Below I provide overall comments on the paper as well as line-by-line comments on the text (many of which are minor issues). Responses to the questions to reviewers are also included at the end of the review.

OUR RESPONSE: Thanks for the positive feedback!

## Overall comments

>  I think the paper would benefit from a brief discussion on how R's handling of factors makes it difficult for new learners. While all the examples you showcase in the paper address important issues, I can see an expert R user suggesting that they have just as robust ways of solving these problems using base R only. However for a new learner a robust base R method is likely to have a much steeper learning curve than the tidyverse method. I think this is a big selling point for the tidyverse method, especially within the context of teaching.

OUR RESPONSE: We've added this point in the introduction. 


>  I suggest dropping the "so-called" in front of tidyverse, unless this is some commonly agreed upon phrasing that I am unaware of.

OUR RESPONSE: We've dropped the so-called. 

> To someone unfamiliar with the tidyverse it might not be obvious that dplyr is part of it. Perhaps defining tidyverse at the beginning of the paper, and listing packages from it that will come up in the paper might be helpful.

OUR RESPONSE: We've added a list of a few packages. 

> Changing the labels of factor levels: Compact but fragile base R example uses a different variable than the robust but verbose base R and Direct and robust dplyr. It would make sense to use the same example for showcasing all three approaches.

OUR RESPONSE: XX AM to address?

I'm struggling with this. I know in other places we have saved the variable as a new object so we don't overwrite anything in the original data, but I like how the repeated dollar signs appear here. This doesn't have the same feeling: 

```{r, eval=FALSE}
laborForce <- GSS$Labor.force.status
levels(laborForce) <- c(levels(laborForce)[1:5], 
                             "Temporarily not working", 
                             "Unemployed, laid off", 
                             "Working full time", 
                             "Working part time")
summary(laborForce)
```

I've edited the examples to all use the laborForce status but do slightly different changes. 


> The defensive coding section could benefit from a discussion on why both assertthat and testthat packages are useful. Is there functionality in one package that is not in the other? If so, the examples should reflect this. If not, and both are showcased just to provide alternatives, it would be a good idea to mention this. Also, I believe the assertions are often used within functions, but the example given here is not within a function. Why would someone not just use "length(levels(drinkstat))==3", which would give TRUE/FALSE, but instead put this within the assert_that function? It would be useful to spell out the reason for the reader.

OUR RESPONSE: XX 

> The tone is a bit too colloquial at times. I think the paper would benefit from reading over once more and making edits specifically to address this.

OUR RESPONSE: We have worked to refine the wording to be more academic while remaining accessible.


## Line-by-line comments

> Line 16: Erronous ? in front of "FitzJohn" in the citation reference

OUR RESPONSE: Fixed

> Line 17-18: "because so many variables are categorical": I understand the sentiment here but this phrasing sounds a bit odd. 

OUR RESPONSE: We've deleted "so".

> Line 45-49: Peng 2015 also mentions an important reason why factors exist in the first place -- usage of categorial variables in regression models. I think it's important to mention this reason here to provide a reason for why one might want to store their categorical variables as factors.

OUR RESPONSE: We've added a phrase about the use of factors in regression models.

> Line 56: "unfortunate behavior of factors in base R": I suggest removing "base" from this phrase as factors behave the same in base R and in tidyverse. The difference is in the syntax of wrangling the factors, but not in how a factor variable behaves.

OUR RESPONSE: Fixed.

> Line 56-60: "online movement": I believe I know what you're referring to here but I find it odd to refer to it as such in a journal article without providing background as to what is meant by an online movement. Also, I think it would be worthwhile to briefly explain what the readr package does (provides functions for importing and exporting data) so that it's clear it is during the import/export stage that strings are kept as is. I would also suggest citing the package here.

OUR RESPONSE: We've cited Peng's paper.  XX NEED TO CITE READR? XX

> Line 61: It's unclear which storage issues have been solved and how they have been solved.

OUR RESPONSE: We've addressed this (see response to Hadley's comments).

> Line 67-68: The ordering of the levels could also be important for interpretability of model results when this variable is an explanatory variable in a multiple linear regression setting.

OUR RESPONSE: Thanks for this, we've changed the modeling section to more explicitly reference linear regression. 

> Line 87: "It had 9 factor levels." = "It is a categorical variable with nine levels."

OUR RESPONSE: We've made this change.

> Line 89-90: "changing the text of one (or more) factor labels, so it appears" = "changing the text of factor labels, so they appear" 

OUR RESPONSE: We've made this change.

> Line 90: This would be relevant in a ggplot2 plot or a base plot. I suggest just referring to a plot, as opposed to specifically a ggplot2 plot.

OUR RESPONSE: We've made this change.

> Line 97: tick after k should be ', not `

OUR RESPONSE: Fixed

> Line 99: Labor.force.status variable should be in monospace font

OUR RESPONSE: Fixed

>  Note: It seems like none of the inline code is formatted properly, e.g.  Line 127. This might be a global setting issue in the Rnw document.

OUR RESPONSE: I think this was related to moving from Rmd to Rnw. Should be fixed now. 

> Line 99: "throw error" - I think the verb "throw" comes from other languages, like C# or JavaScript. I have not heard it used within the context of R, is this common?

OUR RESPONSE: We have replaced the more colloquial "throw an error" with "generate an error message".

> Line 112: "vector function" = "vectorized function"

OUR RESPONSE: We've made this change.

> Line 115-117: Nice example! It would be nice to line number the code chunks so that you can refer to the trailing space with a line number.

OUR RESPONSE: XX

> Line 130: "whitespace an `Below" = "whitespace and `Below" (missing d in and)

OUR RESPONSE: We've made this change.

> Line 132-135 + following code chunk: This example touches upon an important mistake, but it's not only relevant to wrangling categorical data hence I think it should be removed.

OUR RESPONSE: XX

> Line 144: The order is broken the second time this piece of code is run, I don't think it's accurate to say it gets worse. 

OUR RESPONSE: We've changed this to ``the more mixed up the labels and observations get''

> Line 147: Consider removing referring to forcats as a "new addition" since by the time this is published it might not be as new anymore.

OUR RESPONSE: Fixed.

> Line 151: I am not sure calling working in the console an "interactive session" is accurate --- ***COLIN***

OUR RESPONSE: We've elided the mention of the console.

> Line 166: `stringsasfactors=FALSE` = `stringsAsFactors = FALSE` (capitalization and spaces)

OUR RESPONSE: We've made this change.

> Line 167: "final category containing the lumped remainder of people" = "final category that lumps the remainder of people into one" (or some other phrasing, "lumped remainder" sounds awkward)

OUR RESPONSE: We've made this change.

> Line 185: "two category factor" = "two level factor"

OUR RESPONSE: We've made this change.

>Line 187: Would one really write a function for this? If the idea is to have a helper function for such a task I would imagine the number 65 would be an argument for that function. Otherwise, if this is a one time task, I doubt many people would functionalize this task. What was the reason for presenting this task as a function followed by sapply?
	- Also, "x<65" = "x < 65"

OUR RESPONSE: Looking at this code, it's hard to remember why it got written this way! We agree that it's a strange way to solve the problem, and have removed the function and sapply. 


> Line 194-195: I don't follow what this warning is saying, can you clarify? Or maybe give an example?

OUR RESPONSE: XX

> Line 196: "Variable" = "Variables" (plural sounds more appropriate in a section title)

OUR RESPONSE: Fixed

> Line 199 and Line 204: Is there a typo in how the citation is displayed?

OUR RESPONSE: XX

> Line 200: "women, or men" = "women or men"

OUR RESPONSE: We've fixed this.

> Line 205: It sounds awkward to refer to people as "alcohol-involved", is this a technical term?

OUR RESPONSE: We've rewritten the sentence to avoid the use of "alcohol-involved" and instead describe them as having alcohol as their primary substance of abuse.

> Code chunk following Line 212: 
	- "id==1" = "id == 1"
	- Used if_else in an earlier chunk but ifelse here, would be good to keep those consistent (unless there is a reason not to, which I believe there isn't)

OUR RESPONSE: XX

> Code chunk following Line 210: Does mosaic also load mosaicData? If not, it should be loaded as well since it's referred to earlier in the text. 

OUR RESPONSE: XX

> Code chunk following Line 212:
	- "empty repository" = "empty vector"
	- Add whitespaces around logical operators, i.e. "==" = " == "
	- This chunk is quite difficult to follow. I suppose that is the point, so it makes that point very well. But a few comments in especially the moderate section would help readability

OUR RESPONSE: XX
	
> Line 217: "their extreme value" = "the extreme i2 value for that observation" (or something along these lines, current phrasing made me ask "who are they?")

OUR RESPONSE: XX

> Code chunk following Line 218:
	- Note on case_when: I assume this comment will be removed and code chunk updated? (On a related note I really wish case_when worked like this!)
	- Add whitespaces around == for sex variable
	- "this can't be NA": Why not?
	- `tally` is a mosaic function, not dplyr. Since this section is the dplyr section I think it would be more appropriate to use something like HELPsmall %>% count(drink_stat)

OUR RESPONSE: XX
	
> Line 222: What do you mean by "harder to miss possible edge cases"?

OUR RESPONSE: XX

> Code chunk following Line 230: "length(levels(drinkstat))==3" = "length(levels(drinkstat)) == 3" (whitspaces around ==)

OUR RESPONSE: XX

> Code chunk following Line 233: 
	- Will the comments about the code that doesn't work be removed? If not, what is the benefit of keeping them in?
	- Last three lines of code: Explain why all three are necessary, or if not, why they are interchangeable.

OUR RESPONSE: XX

	
> Line 235-236: "Aspects of data wrangling involving categorical variables can be problematic and error-prone." = "Aspects of data wrangling in R involving categorical variables can be problematic and error-prone." (since the discussion in this paper is about doing this in R, not other languages/packages)

OUR RESPONSE: XX

> Line 238: "tidyverse mini-language of R": I like the term "mini-language", but is that commonly used?

OUR RESPONSE: We've dropped the mini-language reference.

## Responses to questions for reviewers

1. Is it useful to demonstrate two ways to do each thing (as long as one isn’t totally stupid) -- Definitely, I like presenting both ways.

2. Do we clarify why each of the tasks are hard? -- I'm not sure what this question means. Do you mean why these tasks are hard due to R's treatment of factors, or why in general data wrangling is hard? Either way, I think the paper clarifies why these tasks are important, but I'm not sure about "hard".

3. Do we clarify why each of the standard approaches are error-prone? -- Yes, especially the reordering factor levels section gives makes a good case about this.

4. Should we focus more on missing values? less? -- I don't think more emphasis on missing values is needed since they are not specific to categorical data.

5. Add appendices or online resources for other examples? Move closing exercise to be online only? -- I suggest providing the GSS data among online resources so that all examples in the paper can be reproduced by the reader.

OUR RESPONSE: XX


6. Add other references? -- The paper indeed is a bit light on references compared to most journal articles. That being said, it doesn't read like it's not grounded in literature, so I don't think references need to be added just for the sake of adding references. 
