---
title: "Response to Reviewers"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Zev review

- We've edited the abstract to be more indicative of the content of the paper, and more references have been incorporated into the introduction for additional context. 

- Removed 'much' from paper, and reduced number of 'that's.

- We agree with the comment about "the complexity of some transformations," so we have removed that and added two additional sentences with high-level descriptions of problems that might occur. 

- To your comment that "all R code" could have new data passed to it and produce updated results, we believe that all *good* R code has these qualities, but it is certainly not true of all R code. For example, analysis which relies on line numbers for subsetting will not perform as expected on updated data. 

- replaced "pejoratively" with "colloquially"

- edited the section on character vectors in linear modeling to be more accurate. 

- changed sentence from "silently fail" to ``the code will not throw errors, but will mislabel all the data such that the `Labor.force.status` variable is essentially meaningless''

- With reference to your comment on **Line 123**, your example actually fails because of a small, hard-to-notice typo. Before your code, the total for Below average was 666, after your code it is 0, because the A in Average is not capitalized. This is the sort of thing that is SO easy to do. I've included a modified version of this example in the paper. 

- Since you didn't actually provide code related to **Line 138**, it's hard to comment directly. However, an approach like this does not work:

```{r}
summary(GSS$Marital.status)
GSS$Marital.status <- factor(GSS$Marital.status, levels = c("Single", "Married", "Single", "No Answer", "Single", "Single", NA))
summary(GSS$Marital.status)
```

- 