---
title: "Working with factor variables in R"
author: "Amelia McNamara"
date: "June 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Factors are the data type that R uses for categorical data. Storing categorical data as a factor variable is more efficient than storing the same data as a set of character strings, because factor variables only store the factor labels once. 

Factor variables are also important when it comes to modeling, because R will know to consider all cases with the same factor level as belonging to the same group. 

So, factors are important. But, they can often be hard to deal with. Because of the way the group numbers are stored separately from the factor labels, it can be easy to overwrite data in such a way that the original data is lost. In this paper, we will consider the best practices for working with factor data. 

To do this, we will consider data from the General Social Survey. 


## Loading the data

We have several options for how to get this data. We could download it in SPSS or Stata formats and use the foreign package to read it in. The GSS download even provides an R file to do the translation for you. Here is the result of that:

```{r}
source('../data/GSS.r')
str(GSS)
```

Obviously, this is less than ideal. Now, all the factor variables are encoded as integers, but their level labels have been lost. We have to look at a codebook to determine if `SEX == 1` indicates male or female. We would rather preserve the integrated level labels. In order to do this, our best option is to download the data as an Excel file and use the `readxl` package to load it. 

```{r}
library(readxl)
GSS <- read_excel("../data/GSS.xls")
str(GSS)
GSS <- GSS[,-14]
```

That's a little better. Now we have preserved the character strings. But, the data is not yet useable in an analysis. 

## Renaming the variables

One problem is that the variable names (while human readable) are full of spaces, so are hard to use. But, we can rename them. The `rename()` function in the `dplyr` package is a good way to do this. 

```{r}
library(dplyr)

GSS <- GSS %>% 
  rename(LaborStatus = `Labor force status`)
```

## Considering some factor variables
Once we have variable names that are easier to work with, we can begin to think about how the data should be cleaned. 
```{r}
GSS <- GSS %>%
  mutate(LaborStatus = factor(LaborStatus))

levels(GSS$LaborStatus) # I wish I had a piece of dplyr code for this
```


## Changing the labels of factors

One action you might want to take is just to change the text of one (or more) of the factor labels, so it appears more nicely formatted in a `ggplot2` plot, for example.

Here is how I do that in base R. Typically, I end up ruining something in the process of doing this, so I *always* start with a summary call, to check after I have done my attempt. 

```{r}
summary(GSS$LaborStatus)
```
```{r}
levels(GSS$LaborStatus) <- c(levels(GSS$LaborStatus)[1:5], "Temporarily not working", "Unemployed, laid off", "Working full time", "Working part time")
summary(GSS$LaborStatus)
```

## Combining several levels into one
This is another common task. Maybe you want fewer coefficients to interpret in your model, or the process that generated the data makes a finer distinction between categories than your research. For whatever the reason, you want to group together levels that are currently separate. 

How I do this in base R:
```{r}
levels(GSS$LaborStatus) <- c("Not employed", "No answer", "Other", "Not employed", "Not employed", "Not employed", "Not employed", "Employed", "Employed")
summary(GSS$LaborStatus)
```

## Combing many categories into one
In this data, age is provided as an integer for respondents 18-88, but then also includes the possible answer "89 or older" as well as a possible "No answer" and NA values. 
```{r}
GSS <- GSS %>%
  mutate(`Age of respondent` = factor(`Age of respondent`))
summary(GSS$`Age of respondent`)
```

We might want to turn this into a factor variable with two levels: 18-65, and over 65. In this case, it would be much easier to deal with a conditional statement about the numeric values, rather than writing out each of the numbers as a character vector.  

## Acknowledgements

Thanks to my students Kelcie Grenier, Kat Kyuchukov, and Emily Ruppel, whose spring 2016 project in my SDS 291 class formed the inspiration for this paper. 
